Redshift Evolution of the Underlying Type Ia Supernova Stretch Distribution

### 2020-09-17 Nicolas.tex
Including MR modifications. Referee answer on GD. YC's comments to be in

\author{
    N. Nicolas \thanks{n.nicolas@ipnl.in2p3.fr, equal contribution} \inst{1} 
    \and M. Rigault \thanks{m.rigault@ipnl.in2p3.fr, equal contribution} \inst{2}
    \orcid{0000-0002-8121-2560}
    \and\\ Y. Copin \inst{1}
    \orcid{0000-0002-5317-7518}
    \and R. Graziani \inst{2}
    \and G. Aldering\inst{3}
    \and M. Briday\inst{1}
    \and J. Nordin\inst{4}
    \and Y.-L. Kim\inst{1}
    \orcid{0000-0002-1031-0796}
    \and Saul Perlmutter\inst{3}
    \and M. Smith\inst{4}
    \orcid{0000-0002-3321-1432}
}

@@

\abstract{The true nature of type Ia supernovae (SNe~Ia) remains largely
    unknown, and as survey statistics increase, the question of astrophysical
    systematic uncertainties rises, notably that of the SN~Ia population
    evolution. In this paper, we study the dependence with redshift of the SN~Ia
    \texttt{SALT2.4} lightcurve stretch, a purely intrinsic SN property, to
    probe its potential redshift drift. The SN stretch has been shown to
    strongly correlate with the SN environment, notably with stellar age
    tracers. We model the underlying stretch distribution as a function of
    redshift, using the evolution of fraction of young and old SNe~Ia as
    predicted by \cite{rigault2018}, and assuming constant underlying stretch
    distribution for each age population made of Gaussian mixtures. We test our
    prediction against published samples chosen to have negligible magnitude
    selection effects, so that any observed change is indeed of astrophysical
    and not observational origin. We clearly demonstrate that the underlying
    SN~Ia stretch distribution is evolving as a function of redshift, and that
    the young/old drifting model is a much better description of the data than
    any time-constant model, including the sample-based asymmetric distributions
    usually used to correct Malmquist bias. The favored underlying stretch model
    is the bimodal one derived from \cite{rigault2018}: a high-stretch mode
    shared by both young and old environments, and a low-stretch mode exclusive
    to old environments. The precise impact of the redshift evolution of the SN
    Ia population intrinsic properties on cosmology remains to be studied. Yet,
    the astrophysical drift of the SN stretch distribution does affect current
    Malmquist bias corrections and thereby distances derived from SN affected by
    selection effects. We highlight that such a bias will increase with surveys
    covering increasingly larger redshift ranges, which is particularly
    important for LSST.}

@@

\subsection{\mr{Testing the construction of a volume limited sample}}\label{ssec:verify}

\mr{In this analysis, we have built a volume limited sample from a set of
magnitude limited ones, using a simple redshift limit.  A ongoing follow up
analysis will make use of the SNANA package \citep{SNANA} to account for
more subtle selection functions. However, the strength of the current
approach is that it is independent of potential inaccuracy in the modeling
of the complex selection functions of the aforementioned dataset. Yet, it is
necessary to test if the sample we have created could show sign of leftover
sample selections.}

\mr{To do so, we compare the stretch and color distributions of the SNe~Ia in
our sample between datasets of overlapping redshifts. If indeed we have built
a volume limited sample, the distributions of SNe~Ia parameters at a given
redshift should not depend of their origin surveys. When considering a
redshift range, this has to be narrow enough such that any drift would be
negligible. The two samples that overlap the most in redshift are PS1 and
SDSS in the redshift range $0.10 < z < 0.20$ (see Fig~\ref{fig:sample}).
This corresponds to the 146 SNe~Ia from SDSS at the highest redshift end and
thus potentially the most affected by potential selection effect leftover
(see the corresponding discussion in section~\ref{sec:sample}). In that same
redshift range, PS1 has 52 SNe~Ia, which are at in the lowest redshift bin
and thus unlikely to have any selection function issues. We present the
stretch and color distributions of both samples in the redshift range $0.10 <
z < 0.20$ in Fig.~\ref{fig:distrib}, where we also provide the p-values of
Kolmogorov-Smirnov tests. The high p-values both in color and stretch
distributions ($>10\%$) strengthen the visual impression that there is no
reason to believe that the SDSS and PS1 SNe~Ia in their overlapping redshift
range are not drawn from the same underlying distributions. A similar
analysis is doable for PS1 and SNLS in their redshift range $0.20 < z <
0.30$ where the same conclusion could be drawn, i.e., that there is no sign
of discrepancy between in the stretch and color distributions between the
low- and high-end of the SNLS and PS1 datasets, respectively. Nonetheless,
the small size of the SNLS dataset at $z < 0.30$ (12 SNe~Ia vs. 90 for PS1)
limits the interest of the test as only strong deviation would be
noticeable. Extending the redshift range to $0.20 < z < 0.40$ (though we have no
PS1 data above 0.3) enables to get 31 SNe~Ia from SNLS in total, the stretch
and color p-values remains high, at 34\% and 4\% respectively.}

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{Article_figures/both-cut_SDSS_PS1-010-020.pdf}
    \caption{\nn{$x_1$ and $c$ normed histograms of the SDSS and PS1 samples for
        $0.10 < z < 0.20$ and their errorbars. On the left, we find in orange
        lines the Base model distributions for $z = 0.10, 0.15, 0.20$ from top
        to bottom. The p-values are the results from a Kolmogorov-Smirnov test;
    they don't show any indication that the samples are not taken from the same
distribution, for both $c$ and $x_1$.}}
    \label{fig:distrib}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{Article_figures/both-cut_SNLS_PS1-020-030.pdf}
    \caption{\mr{Same as Fig.~\ref{fig:distrib} for the PS1 and SNLS data in
    their overlapping redshift range $0.20 < z < 0.30$.}}

    \label{fig:distrib-ps-snls}
\end{figure}

@@

\nn{In order to ensure that the effects we model are only due to the supposed
drift, we need to check for possible Malmquist bias left in the sample.
Because color is most sensitive to selection effects, we studied both the
stretch and color of the end of SDSS and the start of PS1, for $0.10 < z <
0.20$. In this redshift range, }

\nn{The results are shown Fig. \ref{fig:distrib}. We represented the normed
histograms along with their error bars, and found that they don't differ much.
To ascertain this idea, we used a Kolmogorov-Smirnov test that gave us a p-value
of $0.265$ for the color, and $0.137$ for the stretch; while it doesn't tell us
that both the SDSS and PS1 parts come from the same distribution, as would be
expected if we were free from selection effects (so that the samples were random
draws from what Nature could give us), it is not excluded.}

@@

\cite{rigault2018} presented a model for the evolution of the fraction of
younger and older SNe~Ia as a function of redshift following former work on
rates and delay time distributions \citep[e.g.,][]{mannucci2005,
scannapieco2005, sullivan2006, aubourg2008, childress2014, maozmannucci2014}.
In short, it was assumed that the number of ``young'' SNe~Ia follows the star
formation rate (SFR) in the Universe, while the number of ``old'' SNe~Ia
follows the number of Gyr-old stars in the Universe, i.e. the stellar mass
(M$^*$). Hence, if we denote $\delta(z)$ (resp. $\psi(z) = 1-\delta(z)$) the
fraction of young (resp. old) SNe~Ia in the Universe as a function of redshift,
then the ratio $\delta/\psi$ is expected to follow the evolution of the specific
star formation rate (SFR/M$^*$), which goes as $(1+z)^{2.8}$ until $z\sim2$
\citep[e.g.,][]{tasca2015}. Since $\delta(0.05) \sim \psi(0.05)$
\citep{rigault2013,rigault2018,wiseman2020}, in agreement with rate expectations
\citep{mannucci2006,rodney2014}, \cite{rigault2018} concluded that
\begin{equation}
    \label{eq:delta}
    \delta(z) = \left( K^{-1} \times (1+z)^{-2.8} +1 \right)^{-1}
\end{equation}
with $K=0.87$. This model is comparable to the evolution predicted by
\cite{childress2014} based on SN rates in galaxies depending on their quenching
time as a function of their stellar mass.

@@

We also consider another intrinsically non-drifting model, the one developed for
Beams with Bias Correction \cite[BBC,][]{scolnic2016, kessler2017}, used in
recent SN cosmological analyses \cite[e.g.][]{scolnic2018a, descosmopaper2019,
riess2016, riess2019} to account for Malmquist biases. The BBC formalism assumes
sample-based (hence intrinsically non-drifting) asymmetric Gaussian stretch
distributions: $\mathcal{N}\left(\mu,
\sigma_-{}^2\;\text{if}\;x_1<\mu,\;\text{else}\;\sigma_+{}^2\right)$. The idea
behind this sample-based approach is twofold: (1) Malmquist biases are driven by
survey properties and (2) since current surveys cover limited redshift ranges,
having a sample-based approach covers some potential redshift evolution
information \citep{scolnic2016, scolnic2018a}. See further discussion concerning
BBC in Section~\ref{sec:discussion}.

@@

\begin{acknowledgements}
    This project has received funding from the European Research Council (ERC)
    under the European Union's Horizon 2020 research and innovation programme
    (grant agreement n 759194 - USNAC).
    This work was supported in part by the Director, Office of Science, Office
    of High Energy Physics of the U.S.  Department of Energy under Contract No.
    DE-AC025CH11231.
\end{acknowledgements}

@@

% X
% Y
% Z
\end{thebibliography}
\end{document}

>>>>

\author{
    N. Nicolas \thanks{n.nicolas@ipnl.in2p3.fr, equal contribution} \inst{1} 
    \and M. Rigault \thanks{m.rigault@ipnl.in2p3.fr, equal contribution} \inst{2}
    \orcid{0000-0002-8121-2560}
    \and\\ Y. Copin \inst{1}
    \orcid{0000-0002-5317-7518}
    \and R. Graziani \inst{2}
    \and G. Aldering\inst{3}
    \and M. Briday\inst{1}
    \and Y.-L. Kim\inst{1}
    \orcid{0000-0002-1031-0796}
    \and J. Nordin\inst{4}
    \and Saul Perlmutter\inst{3}
    \and M. Smith\inst{4}
    \orcid{0000-0002-3321-1432}
}

@@

\abstract{The true nature of type Ia supernovae (SNe~Ia) remains largely
    unknown, and as survey statistics increase, the question of astrophysical
    systematic uncertainties rises, notably that of the SN~Ia population
    evolution. In this paper, we study the dependence with redshift of the SN~Ia
    \texttt{SALT2.4} lightcurve stretch, a purely intrinsic SN property, to
    probe its potential redshift drift. The SN stretch has been shown to
    strongly correlate with the SN environment, notably with stellar age
    tracers. We model the underlying stretch distribution as a function of
    redshift, using the evolution of the fraction of young and old SNe~Ia as
    predicted by \cite{rigault2018}, and assuming constant underlying stretch
    distribution for each age population made of Gaussian mixtures. We test our
    prediction against published samples chosen to have negligible magnitude
    selection effects, so that any observed change is indeed of astrophysical
    and not observational origin. We clearly demonstrate that the underlying
    SN~Ia stretch distribution is evolving as a function of redshift, and that
    the young/old drifting model is a much better description of the data than
    any time-constant model, including the sample-based asymmetric distributions
    usually used to correct Malmquist bias. The favored underlying stretch model
    is the bimodal one derived from \cite{rigault2018}: a high-stretch mode
    shared by both young and old environments, and a low-stretch mode exclusive
    to old environments. The precise impact of the redshift evolution of the SN
    Ia population intrinsic properties on cosmology remains to be studied. Yet,
    the astrophysical drift of the SN stretch distribution does affect current
    Malmquist bias corrections and thereby distances derived from SN affected by
    selection effects. We highlight that such a bias will increase with surveys
    covering increasingly larger redshift ranges, which is particularly
    important for LSST.}

@@

\subsection{\mr{Testing the construction of a volume limited sample}}
\label{ssec:verify}

\mr{In this analysis, we have built a volume limited sample from a set of
    magnitude limited ones, using a simple redshift limit. An ongoing follow up
    analysis will make use of the SNANA package \citep{SNANA} to account for
    more subtle selection functions. However, the strength of the current
    approach is that it is independent of potential inaccuracy in the modeling
    of the complex selection functions of the aforementioned dataset. Yet, it is
    necessary to test if the sample we have created could show sign of leftover
sample selections.}

\mr{To do so, we compare the stretch and color distributions of the SNe~Ia in
    our sample between datasets of overlapping redshifts. If indeed we have
    built a volume limited sample, the distributions of any SNe~Ia parameter at
    a given redshift should not depend of their origin surveys. When considering
    a redshift range, this has to be narrow enough such that any drift would be
    negligible. The two samples that overlap the most in redshift are PS1 and
    SDSS in the redshift range $0.10 < z < 0.20$ (see Fig~\ref{fig:sample}).
    This corresponds to the 146 SNe~Ia from SDSS at the highest redshift end and
    thus potentially the most affected by potential selection effect leftover
    (see the corresponding discussion in section~\ref{sec:sample}). In that same
    redshift range, PS1 has 52 SNe~Ia, which are at in the lowest redshift bin
    and thus unlikely to have any selection issue. The Fig.~\ref{fig:distrib}
    present the stretch and color distribution of both, together with p-values
    of Kolmogorov-Smirnov tests that test if the $x_1$ and $c$ of the SDSS and
    PS1 samples at $0.10<z<0.20$ arise from the same underlying distribution.
    Those p-values are high ($>10\%$), and if one cannot conclude from this test
    that they SNe~Ia are from the same underlying distribution, one has no
    ground to claim they are not. This is in agreement with the visual
    impression from Fig.  ~\ref{fig:distrib}.  A similar analysis is doable for
    PS1 and SNLS in their redshift range $0.20 < z < 0.30$ where the same
    conclusion could be drawn, i.e., that there is no sign of discrepancy
    between in the stretch and color distributions between the low- and high-end
    of the SNLS and PS1 datasets, respectively. Nonetheless, the small size of
    the SNLS dataset at $z < 0.30$ (12 SNe~Ia vs. 90 for PS1) limits the
    interest of the test as only strong deviation would be noticeable. Extending
the redshift range to $0.20 < z < 0.40$ (though we have no PS1 data above 0.3)
enables to get 31 SNe~Ia from SNLS in total, the stretch and color p-values
remains high, at 34\% and 4\% respectively.}

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{Article_figures/both-cut_SDSS_SNLS_PS1.pdf}
    \caption{\mr{$x_1$ and $c$ distribution histograms of different surveys
    overlapping in redshift. \emph{Facing up} SDSS and PS1 within the redshift
    range $0.10 < z < 0.20$ ; \emph{facing down} PS1 and SNLS within the
    redshift range $0.20<z<0.3$. Error bars show the Poisson noise. Our stretch
    "Base" modeling is illustrated in orange at the mean redshift of the
    redshift ranges, $0.15$ and $0.25$, respectively. P-values of the
    Kolmogorov-Smirnov tests are indicated on the top (bottom) of each panel
    showing no sign that the SDSS and PS1 (PS1 and SNLS) $x_1$ and $c$
    distributions are not drawn from the same underlying distributions.}}
    \label{fig:distrib}
\end{figure}

@@

@@

\cite{rigault2018} presented a model for the evolution of the fraction of
younger and older SNe~Ia as a function of redshift following former work on
rates and delay time distributions \citep[e.g.,][]{mannucci2005,
scannapieco2005, sullivan2006, aubourg2008, childress2014, maozmannucci2014}.
In short, it was assumed that the number of ``young'' SNe~Ia follows the star
formation rate (SFR) in the Universe, while the number of ``old'' SNe~Ia follows
the number of Gyr-old stars in the Universe, i.e. the stellar mass (M$^*$).
Hence, if we denote $\delta(z)$ (resp. $\psi(z) = 1-\delta(z)$) the fraction of
young (resp. old) SNe~Ia in the Universe as a function of redshift, then the
ratio $\delta/\psi$ is expected to follow the evolution of the specific star
formation rate (SFR/M$^*$), which goes as $(1+z)^{2.8}$ until $z\sim2$
\citep[e.g.,][]{tasca2015}. Since $\delta(0.05) \sim \psi(0.05)$
\citep{rigault2013, rigault2018, wiseman2020}, in agreement with rate
expectations \citep{mannucci2006, rodney2014}, \cite{rigault2018} concluded that

\begin{equation}
    \label{eq:delta}
    \delta(z) = \left( K^{-1} \times (1+z)^{-2.8} +1 \right)^{-1}
\end{equation}
with $K=0.87$. This model is comparable to the evolution predicted by
\cite{childress2014} based on SN rates in galaxies depending on their quenching
time as a function of their stellar mass.

@@

We also consider another intrinsically non-drifting model, \mr{the functional
    form} developed for Beams with Bias Correction \cite[BBC,][]{scolnic2016,
    kessler2017}, used in recent SN cosmological analyses
    \cite[e.g.][]{scolnic2018a, descosmopaper2019, riess2016, riess2019} to
    account for Malmquist biases. The BBC formalism assumes sample-based (hence
    intrinsically non-drifting) asymmetric Gaussian stretch distributions:
    $\mathcal{N}\left(\mu,
    \sigma_-{}^2\;\text{if}\;x_1<\mu,\;\text{else}\;\sigma_+{}^2\right)$. The
    idea behind this sample-based approach is twofold: (1) Malmquist biases are
    driven by survey properties and (2) since current surveys cover limited
    redshift ranges, having a sample-based approach covers some potential
    redshift evolution information \citep{scolnic2016, scolnic2018a}. See
    further discussion concerning BBC in Section~\ref{sec:discussion}.

@@

\begin{acknowledgements}
    This project has received funding from the European Research Council (ERC)
    under the European Union's Horizon 2020 research and innovation programme
    (grant agreement n 759194 - USNAC).
    This work was supported in part by the Director, Office of Science, Office
    of High Energy Physics of the U.S. Department of Energy under Contract No.
    DE-AC025CH11231.
    This project is partly financially supported by Region Rhone-Alpes-Auvergne
\end{acknowledgements}

@@

% X
% Y
% Z
\end{thebibliography}
\end{document}
end{document}
